\documentclass[a4paper]{MBE}
\usepackage[bookmarks]{hyperref}
\hypersetup{
	colorlinks,
	citecolor=black,
	filecolor=black,
	linkcolor=black,
	urlcolor=black
}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{lmodern}
\usepackage{bm,bbold}
\usepackage{xfrac, nicefrac}
\usepackage{graphicx}
\graphicspath{{artworks/}}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows,automata}
\pdfinclusioncopyfonts=1

\include{notations}

\jshort{mst}
\volname{}
\jvolume{0}
\jvol{}
\jissue{0}
\pubyear{2019}

\mstype{Article}

\begin{document}
\title[Substitution rate susceptibility] {Substitution rate response to changes in effective population size and protein expression level.}

\author[Latrille \& Lartillot] {
	T. \surname{Latrille}$^{\ast,1,2}$,
	N. \surname{Lartillot}$^{1}$
}

\address{
	$^{1}$Université de Lyon, Université Lyon 1, CNRS, Laboratoire de Biométrie et Biologie Évolutive UMR 5558, 69622 Villeurbanne, France.\\
	$^{2}$École normale supérieure de Lyon, Université de Lyon, Université Lyon 1, Lyon, France}

\coresp{thibault.latrille@ens-lyon.org or nicolas.lartillot@univ-lyon1.fr}


\abstract{
	Molecular sequences are shaped by selection, where the strength of selection relative to drift is arbitrated by effective population size ($\Ne$). Populations with high $\Ne$ are expected to undergo stronger purifying selection, and consequently a decrease in substitution rate of selected mutations relative to the substitution rate of neutral mutation ($\omega$=dN/dS). However, computational models based on biophysics of protein stability have suggested that $\omega$ can also be independent of $\Ne$, proven under general conditions. Together, the response of $\omega$ to changes in $\Ne$ depends on the specific mapping from sequence to fitness. Importantly, increase in protein expression level has been found empirically to result in decrease of $\omega$, predicted by theoretical results recruiting the selection for protein stability. In this present work we derive a theoretical approximation for the response of $\omega$ to changes in $\Ne$ and expression level, under an explicit genotype-phenotype-fitness map. The method is generally valid for additive traits and log-concave fitness functions. We applied the results to protein undergoing selection for their conformational stability, corroborated by simulations under more complex models. We predict a weak response of $\omega$ to changes in either $\Ne$ or expression level, which are interchangeable. Based on empirical data, we propose that fitness based on the conformational stability is not a sufficient mechanism to explain the amplitude of $\omega$ variations observed. However, other aspects of protein biophysics might be explored such as protein-protein interactions, which can lead to stronger response of $\omega$ to changes in $\Ne$.
}

\keyword{protein stability, substitution rate, population-genetics, drift, expression level.}

\maketitle

\include{GenoPhenoFit-main}


\section{Supplementary Material}
The mathematical development of the general case and supplementary figures are compiled in a single file.
The simulators written in C++ are publicly available under MIT license at \url{https://github.com/ThibaultLatrille/SimuEvol}.
The scripts and instructions necessary to reproduce this study are available at \url{https://github.com/ThibaultLatrille/GenotypePhenotypeFitness}.

\section{Acknowledgments}

The authors gratefully acknowledge the help of Nicolas Rodrigue and Laurent Duret for their advice and review concerning this manuscript.
This work was performed using the computing facilities of the CC LBBE/PRABI.
A PhD grant of the École normale supérieure de Lyon awarded to T.L. provided a salary during part of the work.

\bibliographystyle{natbib}%%%%Bibliography style file
\bibliography{references/references}

\end{document}