\documentclass[a4paper]{MBE}
\usepackage[bookmarks]{hyperref}
\usepackage{float}
\usepackage{adjustbox}
\usepackage{tabu}
\usepackage{hhline}
\hypersetup{
	colorlinks,
	citecolor=black,
	filecolor=black,
	linkcolor=black,
	urlcolor=black
}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{lmodern}
\usepackage{bm,bbold}
\usepackage{xfrac, nicefrac}
\usepackage{graphicx}
\graphicspath{{artworks/}}
\makeatletter
\def\input@path{{artworks/}}
\makeatother
\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows,automata}
\pdfinclusioncopyfonts=1

\include{notations}

\jshort{mst}
\volname{}
\jvolume{0}
\jvol{}
\jissue{0}
\pubyear{2019}

\mstype{Article}

\begin{document}
\title[Inferring long-term population size] {Inferring long-term demographic history with Mutation-Selection models.}


\author[Latrille et al.] {
	T.
\surname{Latrille}$^{\ast,1,2}$,
	V.
\surname{Lanore}$^{1}$,
	N.
\surname{Lartillot}$^{1}$}

\address{
	$^{1}$Université de Lyon, Université Lyon 1, CNRS, Laboratoire de Biométrie et Biologie Évolutive UMR 5558, F-69622 Villeurbanne, France.\\
	$^{2}$École normale supérieure de Lyon, Université de Lyon, Université Lyon 1, Lyon, France}


% \history{Received ----; reviews returned -----;}

\coresp{E-mail: thibault.latrille@ens-lyon.org}

% \datade{The data were obtained from OrthoMam}
% \editor{}

\abstract{
	Selection in protein-coding sequences can be detected based on multiple sequence alignments using phylogenetic codon models.
	Mechanistic approaches, grounded on population-genetics first principles, explicitly formalize the interplay between mutation, selection and random drift, and return an estimate of the amino-acid fitness landscape.
	However, these recently developed models rely on the assumption of constant effective population size.
	We propose an extended mutation-selection model reconstructing site-dependent fitness landscape, long-term trends in effective population size and mutation rate along the phylogeny, from an alignment of DNA coding sequences.
	Independently, ancestral life-history traits are reconstructed along the phylogeny from observation in present-day species.
	Together, we estimate the correlation between reconstructed life-history traits, mutation rate and effective population size, intrinsically including phylogenetic inertia.
	Our framework has been tested against simulated data, and in empirical data in mammals and primates.
	Finally, our work also points to important theoretical questions about how coding sequences respond to changes in effective population size and to fluctuating selection.
}

\keyword{Phylogenetic, codon models, mutation-selection models, population genetic, population size, mutation rate, life history traits.}

\maketitle

\include{MutSelDrift-main}

\section{Supplementary Material}
Supplementary materials deriving mathematical formulas and figures are available at "!!!!TO BE WRITTEN!!!!".
Mammalian alignments and life-history traits, simulated alignments and results of experiments are available at \url{https://datadryad.org/"!!!!TO BE WRITTEN!!!!"}.
The scripts and instructions necessary to reproduce the experiments are available at \url{https://github.com/ThibaultLatrille/MutationSelectionDrift}.

\section{Acknowledgments}
The authors gratefully acknowledge the help of Nicolas Rodrigue, Laurent Guegen, Philippe Veber for their advice and review concerning this manuscript.
This work was performed using the computing facilities of the CC LBBE/PRABI.
A PhD grant of the École normale supérieure de Lyon awarded to T.L.
provided a salary during the work.

\bibliographystyle{natbib}%%%%Bibliography style file
\bibliography{references/references}

\end{document}