\chapter*{Preamble}
\addcontentsline{toc}{part}{Preamble}
The diversity of living organisms today is the result of a complex and intricate process, which operates at multiple levels.
At the molecular level, the fate of a protein depends on its ability to fold but also on the enzymes it encounters from its creation up to its degradation.
Composed of billions of proteins, a cell's own fate depends on its own ability to metabolize substrates and copy its \acrshort{DNA}, but also on the fate of surrounding cells and the individual to which it belongs.
Moreover, the fate of this individual depends on its own behaviour, but also on its environment and the population to which it belongs.
Altogether, scientists have dissected this intricate process into its core components, through molecular biology, enzymology, metabolism, physiology, population genetics, ecology, and so on.
Molecular evolution seeks to encompass different levels, relating molecular changes to higher-level evolutionary processes.
In this vein, this work is a modest attempt to reconcile several layers of evolution, mechanistically deriving how observable parameters between populations and within populations depends in microscopic molecular and cellular parameters.
Altogether, through the framework of population genetics, I seek to draw connections between independent datasets, from molecular parameters of protein biophysics, to diversity and divergence of \acrshort{DNA} sequences within and between species, while relating to species' quantitative life-history traits.\\

This thesis is submitted in partial fulfilment of the requirements for the degree of \emph{Philosophiae Doctor} at the Université de Lyon.
The research presented here was conducted at the Laboratoire de Biométrie et Biologie Evolutive (LBBE), under the supervision of research director M. Nicolas Lartillot.
This work was conducted from September 2017 onward during a 3-year grant by ENS de Lyon (Contrat Doctoral Spécifique Normalien).
The thesis is a collection of three manuscripts preceded by an introduction that relates them and provides background information and motivation for the work.

All figures, scripts and \LaTeX source code used in this manuscript can be reused under CC-BY-SA license, available at \url{https://github.com/ThibaultLatrille/PhD}.
